The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-02-17 src/ms.tex
draft draft section on interoperability
Signed-off-by: nstarman <nstarkman@protonmail.com>

The relationship between \astropy and community is reciprocal, with the growing and evolving needs of the community and the maturation of the broader scientific \python ecosystem driving the development of the \astropypkg core package since \citetalias{astropy:2018}.
% feels like ↑ could be expanded to 2 sentences

% Maybe briefly outline the sectors of  
% Using \astropypkg for
% precision astronomy has spurred development of faster and more accurate coordinate transformations and time calculations,
% super-computer work has spurred \astropypkg to be thread-safe and support \package{dask} \citep{daskpkg}
% Astropy has grown to interoperate, e.g. NumPy

Development may be roughly split into a few categories: new features added, performance enhancements, intra and inter-package interoperability, and improvements to precision, accuracy, and reproducibility.

Before discussing those, we should mention that sometimes one realizes
features do not actually belong in the core package, either because
they are of more general use outside of \astropypkg, or because they
are too specialized.  An example of the former is the copy of the IAU
Standards Of Fundamental Astronomy (SOFA) software
Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
astropy carried, as well as the python wrappers for it.  Since this is
more basic infrastructure, with a release cycle set by SOFA, it made
more sense to create separate packages that \astropypkg\ can depend
on: \package{ERFA} \citep{erfa} and \package{PyERFA} \citep{pyerfa}.
An example of something that was too specialized was a Virtual
Observatory sub-package (now removed).  Of its content, the Simple
Application Message Protocol made sense to keep
(\astropysubpkg{samp}), but other parts belonged more properly in {\tt
  Astroquery} (see Sect.~\ref{sec:astroquery}).

% Summarize major pieces of development since the last paper (since v2.0). A few
% ideas (taken from What's New pages) below, but feel free to expand on or remove
% from this list:
% \begin{itemize}
%     \item Support for representing and transforming velocity data in coordinates, epoch propagation (v3.0)
%     \item Improved support for astronomical time series: TimeSeries object (v3.2), Box Least Squares periodogram (v3.1)
%     \item Overall improved support of Quantity throughout numpy (v4.0) and scipy
%     \item Native support for Time, Quantity, and SkyCoord objects in Astropy tables
%     \item APE 14 WCS
%     \item A new SpectralCoord class for representing and transforming spectral quantities
%     \item Support for dask/large arrays in several places in the core package (4.1, 5.0, and maybe other versions)
% \end{itemize}

@@

@@

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-performance}

\subsection*{Precision, Accuracy, \& Reproducibility} \label{sec:core-features-future}

>>>>

  The relationship between \astropy and community is reciprocal, with the
  growing and evolving needs of the community and the maturation of the broader
  scientific \python ecosystem now driving the development of the \astropypkg
  core package.

  One manifestation of this development realizing some features do not actually
  belong in the core package, either because they are of more general use
  outside of \astropypkg, or because they are too specialized. An example of the
  former is the copy of the IAU Standards Of Fundamental Astronomy (SOFA)
  software Collection\footnote{\url{http://www.iausofa.org}} \citep{sofa} that
  \astropy carried, as well as the \python wrappers for it.  Since this is more
  basic infrastructure, with a release cycle set by SOFA, it made more sense to
  create separate packages that \astropypkg can depend on: \package{ERFA}
  \citep{erfa} and \package{PyERFA} \citep{pyerfa}. An example of something that
  was too specialized was a Virtual Observatory sub-package (now removed).  Of
  its content, the Simple Application Message Protocol made sense to keep
  (\astropysubpkg{samp}), but other parts belonged more properly in {\tt
  Astroquery} (see Sect.~\ref{sec:astroquery}).

  Within \astropypkg, development may be roughly split into a few categories:
  new features added, performance enhancements, intra and inter-package
  interoperability, and improvements to precision, accuracy, and
  reproducibility. We discuss these in turn.

@@

  % New features capture a lot of attention. While exciting, the arguably more
  % significant improvements are those of interoperability -- ensuring existing
  % features connect smoothly together.

  The modular nature of \astropypkg means that not all sub-packages deeply
  integrate with one another -- and similarly between separate \python
  libraries. Because \astropy is situated at the nexus between scientific
  computing into astronomy, interoperability has been and continues to be an
  important focus for \astropy, both within the various \astropypkg sub-packages
  and with other important \python libraries. While increasing inter and
  intra-package integration will be a perpetual goal -- with feature additions
  in one place requiring updates in another -- \astropy has made impactful
  progress with units, tables, and I/O.
  % Better list? Also better transition
  If done right, all a user notices is that code using many different python
  features and libraries ``just works''. Here we will draw notice to these
  efforts to make your code ``just work''.

  \paragraph{\package{numpy} on Units}

    \astropyQuantity\ is a backbone of \astropy, leveraging the power of
    \package{numpy} and adding units. Previously, many \package{numpy} functions
    would strip a \astropyQuantity\ of its units (or fail outright), limiting
    \astropyQuantity's usefulness. Now, advances in \package{numpy} function
    overloading (see for example
    \href{https://numpy.org/neps/nep-0018-array-function-protocol.html}{NEP 18})
    mean \astropyQuantity\ works with almost all \package{numpy} (version 1.7+)
    functions. These advances have similarly benefitted many
    \package{numpy}-dependent packages, like \package{scipy} and their sparse
    matrices. And again \astropypkg benefits since more \package{scipy}
    functions now work with \astropyQuantity.

    In the remaining gaps, the \package{numpy} and \astropyQuantity\
    interoperability efforts are ongoing. For some functions, such as in
    \href{https://numpy.org/doc/stable/user/basics.rec.html#module-numpy.lib.recfunctions}{\package{numpy}'s
    module} for manipulating structured arrays, compatibility only requires
    exending the existing \package{numpy}-\astropypkg bridge framework.
    Community interest, in the form of a Feature Request (or better yet Pull
    Request) would be sufficient to see this realized. For other functions,
    discussed further in the
    \href{https://docs.astropy.org/en/stable/known_issues.html#known-deficiencies}{\astropy
    documentation}, the \package{numpy} framework does not yet allow for full
    interoperability with \astropy. A goal of \astropy, and the scientific
    \python community is to enhance and implement the frameworks, allowing
    \astropyQuantity\ to propagate units seamlessly across the whole ecosystem of
    \package{numpy}-like projects.

  \paragraph{Astropy in Time and on Units}

    % Bad sentence ↓. Rewrite.
    Not only did \astropyunits gain powerful integrations with \package{numpy},
    there were many intra-package improvements.

    \astropymodeling can add units to otherwise unit-less models.

    A new module –
    \astropycosmologyunits
    – has been added to the cosmology sub-package for defining and collecting
    cosmological units and equivalencies. The unit \texttt{littleh} and
    equivalency
    \astropyapidoc{cosmology.units.with_H0}{with\_H0}
    are deprecated from the main \astropyunits sub-package and moved to
    \astropycosmologyunits. A new unit, \texttt{redshift}, is added for tracking
    factors of cosmological redshift.
    % As this is a pseudo-unit an equivalency
    % \astropyapidoc{cosmology.units.dimensionless_redshift}{dimensionless\_redshift}
    % is added (and enabled by default) to allow for redshift - dimensionless
    % conversions.
    To convert between redshift and other cosmological distance
    measures, e.g. CMB temperature or comoving distance, the equivalency
    \astropyapidoc{cosmology.units.with_redshift}{with\_redshift}
    is also added. This equivalency is actually a composite of other
    equivalencies,
    % : \astropyapidoc{cosmology.units.redshift_distance}{redshift\_distance},
    % \astropyapidoc{cosmology.units.redshift_hubble}{redshift\_hubble}, and
    % \astropyapidoc{cosmology.units.redshift_temperature}{redshift\_temperature},
    which may be used separately.

    % IN PROGRESS (ADAPTING FROM THE WHATSNEW).
    The FITS Time standard paper defines the formats and keywords used to
    represent timing information in FITS files. The Astropy FITS package now
    provides support for reading and writing native Time table columns using
    this format. This is described within the FITS unified I/O interface and
    examples of usage can be found in the Astropy native objects (mixin columns)
    section.

    Implementation of this feature was made possible through the support of the
    Google Summer of Code 2017 program. To learn more about this project please
    see the final report, A mixin protocol for seamless interoperability.

  \paragraph{Table Mixin Columns}

    Numerous \python libraries have custom table implementations
    \begin{itemize}
      \item (Support use of SkyCoord as Column happened before) and in table vstack, dstack, and insert\_row
      \item Converting SkyCoord to QTable (1 sentence addition to above)
      \item Added support for dask arrays in tables
      \item Added support for registering array-like objects as mixin columns
    \end{itemize}

  \paragraph{Unified I/O architecture}

    \astropyiosubpkg{registry} is a powerful way to define input and output
    (I/O) functions for \astropypkg objects, such as reading or writing to a
    file. \astropy uses this internally for the I/O methods in \astropyTable and
    \astropyCosmology (I/O added v5.0+), and users can register custom I/O to
    extend the options on these classes.

    As of Astropy 5.0 the I/O registry submodule has switched to a class-based
    architecture and been refactored to allow read-only, write-only, and
    read-write registries. Additionally, the registry submodule no longer
    assumes an object has \astropy-style \texttt{read} and \texttt{write}
    methods when trying to read or write that object. These two improvements
    enable a number of new use cases for \astropyiosubpkg{registry}, some
    highlighted here, but others we hope to see have community adoption.

    First, the class-based architecture allows for the creation of custom
    registries. An application of this is in \astropycosmology, which
    has two different \textit{kinds} of registries for \astropyCosmology: one
    for reading and writing files, and another for converting between python
    objects.

    Second, \textit{any} kind of object is now compatible with
    \astropyiosubpkg{registry}. Using \astropyiosubpkg{registry} for custom
    classes means a user does not have to create a whole new I/O system and can
    offer a unified and familiar API.

  \paragraph{Persistent Storage and I/O}

    % TODO Better transition sentence
    The imrovements to \astropy's I/O architecture are mirrored in the increased support for I/O options.

    \begin{itemize}
      \item ASDF : JSWST  and models can be stored with ASDF.
      \item ECSV : Support for multidimensional and object columns in ECSV
      \item FITS : Support for writing Dask arrays to FITS files, and Append table to existing FITS file
      \item Support for reading, writing, and converting Cosmology
      \item MRT Support for reading and writing tables to MRT format : GSOC
      \item Support for reading and writing tables to Parquet format
      \item Support for reading and writing tables to QDP format
      \item ASDF read write
    \end{itemize}

@@

  An important focus of \astropy has been to increase the precision and accuracy
  of its functionality, while, where appropriate, making sure results obtained
  using older versions of \astropypkg are reproducible.
  % Something about ScienceState
  Notable improvements have been made to \astropytime, \astropyconstants,
  \astropycoordinates, and \astropycosmology.

  \paragraph{Time}

    In astronomy, time accuracy down to the second is frequently important.
    Whether for planning observations or crunching data, missing seconds can
    meaningfully impact results. Before, when using \astropyTime in \astropytime
    leap seconds had to be manually applied; now leap seconds are applied
    automatically. Moreover, \astropypkg updates internal time data-files to
    ensure that the correct leap second adjustment is always used. For
    reproducibility, e.g. with an old code, \astropyLeapSeconds may also be
    manually applied.

  \paragraph{Constants}

    Measurements of physical constants (and the units defined from them) improve
    over time. Periodically the standardized systems of units and constants are
    updated to reflect these improvements, for instance in 2019 the SI system was
    redefined \citep{NIST2019}, with an accompanying update to the physical
    constants in
    \href{https://codata.org/initiatives/data-science-and-stewardship/fundamental-physical-constants/}{\texttt{SI/CODATA
    2018}}. \astropypkg contains two modules -- \astropyconstants and
    \astropyunits -- for working with constants and units, respectively.
    \astropyconstants now defaults to use the \texttt{SI/CODATA 2018} values, with
    the relation between units in \astropyunits based upon these constants. Most
    of \astropypkg and affiliate packages build upon \astropyunits, so this update
    has far-reaching effects across the \astropy ecosystem and astronomy
    community.

    For reproducibility, \astropypkg allows the constants (and therefore units)
    definitions to be rolled back to prior values, for instance the
    \texttt{SI/CODATA 2014} definitions.
    % CITATION
    The constants and units definitions are set by configuration file as well as
    \astropyScienceState (an runtime configuration controller). For work where
    updates to the values of the fundamental constants significantly alters
    results we recommend including with the work an astropy configuration file,
    which specifies the set of constants used.

  \paragraph{Cosmology}
    % TODO! At least one good sentence about "what is the benefit of astropy.cosmology"

    \astropycosmology contains classes for representing cosmological models.
    % currently all homogeneous and isotropic variations of the
    % Friedmann-Lemaitre-Robertson-Walker metric.
    Bundled with the classes are commonly used \astropyCosmology realizations, for
    example best-fit measurements from
    \href{https://lambda.gsfc.nasa.gov/product/map/current/}{WMAP}
    \citep{WMAP2003} and \href{https://www.nasa.gov/mission_pages/planck}{Planck}
    \citep{PlanckMission:2006}.

    When these important missions publish new measurements, \astropycosmology is
    updated to include these results as realizations of the appropriate class
    (generally \astropyFlatLambdaCDM). The models used by WMAP and Planck do not
    always exactly correspond to \astropycosmology classes, for instance the
    Planck 2018 results \citep{Planck2018VI} include massive neutrinos in
    $\Omega_{matter,0}$, while in \astropyFlatLambdaCDM this mass contribution is
    stored in a separate parameter. Consequently, while some parameter values
    appear different from the source paper, the \astropyCosmology realizations
    correctly reproduce the cosmological models.

    Since \citealt{astropy:2018} the following cosmology realizations have been
    added:
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP1.html}{\texttt{WMAP1}}
    from WMAP First Year \citep[Table 7: WMAP + CBI + ACBAR + 2dFGRS +
    Lya]{WMAP1Year:2003};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.WMAP3.html}{\texttt{WMAP3}}
    from WMAP Three Year \citep[Table 6: WMAP + SNGold]{WMAP3Year:2007};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck15.html}{\texttt{Planck15}}
    from Planck 2015 \citep[Table 4: TT, TE, EE + lowP + lensing +
    ext]{Planck2015XIII:2016};
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.Planck18.html}{\texttt{Planck18}}
    from the Planck 2018 best-fit cosmological parameters \citep[Table 2: TT, TE,
    EE + lowE + lensing + BAO]{Planck2018VI:2020}.

    Astropy provides a
    \href{http://docs.astropy.org/en/stable/api/astropy.cosmology.default_cosmology.html}{configurable
    default cosmology}. All calculations done in a cosmological context will use
    this default cosmology. The default \astropyCosmology has been updated to the
    Planck 2018 parameters. For reproducibility, the default cosmology may be set
    to old defaults, such as the WMAP Three Year values. The default cosmology is
    dynamically configurable using \astropyScienceState, meaning a set of
    calculations may be run using different assumed cosmologies, and results
    compared between the two.
    % Repeat the word default less.

  \paragraph{Coordinates}

    Being able to describe the position of objects, i.e. their coordinates, is
    fundamental across many knowledge domains. \astropypkg provides
    \astropycoordinates for working with low-level positional and velocity data
    all the way to high-level coordinate objects with reference frames,
    atmospheric information, and more. A central feature of \astropycoordinates is
    the ability to transform data between reference frames, e.g. \astropyICRS
    \citep{ICRS:1997} to \astropyGalacticLSR \citep{GalacticLSR:2010}. However,
    there have been some historical limitations for spatially proximate
    transformations, impacting the usefulness of \astropycoordinates for
    ground-based telescopes and intra-solar astronomy. Transformations in the
    \astropyAltAz frame were reasonably precise for very distant objects, but
    wrong by up around several arcseconds for e.g. the location of the moon. Now
    these transformations are much more precise, down to the milli-arcsecond
    level. Similar precision improvements were made to the
    \astropyapi{coordinates.builtin_frames.HADec}{Hour Angle-Declination} frame
    transformations. Additionally,
    \astropyapi{coordinates.builtin_frames.BaseEclipticFrame}{Ecliptic frames}
    and associated transformations have been updated to correctly reflect the
    “true” (includes precession but not nutation) and “mean” (precession and
    nutation) terminology.

    For Galactic astronomers, the \astropyGalactocentric
    \astropyapi{coordinates.galactocentric_frame_defaults}{frame defaults} have
    been updated to include more recent measurements (tabulated in the
    \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html#astropy.coordinates.galactocentric_frame_defaults.references}{\texttt{frame\_defaults.references}}
    attribute). For reproducibility, old definitions are still available, and will
    continue to be so as \astropypkg is updated with forthcoming measurements of
    the Galactic center, solar motion, etc.
    % may be used by setting the default value to ``pre-v4.0'' in the
    % \href{https://docs.astropy.org/en/stable/api/astropy.coordinates.galactocentric_frame_defaults.html}{\texttt{galactocentric\_frame\_defaults}}
    % \astropyScienceState.

\subsection*{Performance} \label{sec:core-features-performance}

\subsection*{Future} \label{sec:core-features-future}
